\documentclass{article}
\input{preamble}
\input{preamble-biblatex}
\input{preamble-theorems}
\input{preamble-typing}
\input{preamble-emoji}
\pagenumbering{gobble}
\usepackage[section]{placeins}
\usepackage[parfill]{parskip}
\usepackage{titlesec}
\newcommand{\sectionbreak}{\ifnum\value{section}>1\clearpage\fi}
\addbibresource{main.bib}
\author{Wen Kokke}
\title{Hypersequent Classical Processes}
\begin{document}


\section{Classical Processes}\label{sec:cp}
In this section, we will introduce a variant of \cp which removes the
commutative conversions from the reductions.
\cp was introduced by \textcite{wadler2012}, based on work by \textcite{caires2010}
and \textcite{gay2009}.

\subsection{Terms}\label{sec:cp-terms}
\input{def-cp-terms}
\input{def-cp-equiv}
\input{def-cp-reduction}

\subsection{Types}\label{sec:cp-types}
\input{def-cp-types}
\input{def-cp-negation}
\input{lem-cp-negation-involutive}
\input{def-cp-envs}
\input{def-cp-typing}

\subsection{Properties}\label{sec:cp-properties}

\subsubsection{Preservation}\label{sec:cp-preservation}
\input{lem-cp-preservation-equiv}
\input{lem-cp-preservation-subst}
\input{thm-cp-preservation}

\subsubsection{Canonical Forms and Progress}\label{sec:cp-canonical-forms-and-progress}

\paragraph{Canonical Forms}\label{sec:cp-canonical-forms}
\input{def-cp-action}

\paragraph{Evaluation Contexts}\label{sec:cp-evaluation-contexts}
\input{def-cp-evaluation-contexts}
\input{def-cp-evaluation-prefixes}
\input{def-cp-maximum-evaluation-prefixes}
\input{lem-cp-maximum-evaluation-prefixes}

\paragraph{Progress}\label{sec:cp-progress}
\input{thm-cp-progress}


\section{Hypersequent Classical Processes}\label{sec:hcp}
In this section, we will introduce a variant of \cp based on hypersequent
calculi.
Hypersequent calculi were introduced independently by \textcite{avron1987} and
\textcite{pottinger1983}.

\subsection{Terms}\label{sec:hcp-terms}
\input{def-hcp-terms}
\input{def-hcp-equiv}
\input{def-hcp-reduction}

\subsection{Types}\label{sec:hcp-types}
We keep the types and environments from \cp, see \cref{def:cp-types} and
\cref{def:cp-envs}.
We introduce a new layer on top of sequents: hypersequents.
However, as \cp is a one-sided logic, and it uses the left-hand side of the
turnstile to write the process, the traditional hypersequent notation can look
awkward:
\[
  \seq[P]{\Gamma_1}\hsep\dots\hsep\seq{\Gamma_n}.
\]
The above may be confusing. It seems to claim that $\tm{P}$ acts according to
protocol $\ty{\Gamma_1}$. What are all the other $\Gamma$s doing there?
Are they typing empty processes?
Therefore, we opt to leave out the repeated turnstile, and instead work with the
notion of ``hyper-environments''. However, we will still refer to our system as a
hypersequent system.
\input{def-hcp-hyperenvs}
\input{def-hcp-typing}

\subsection{Properties}\label{sec:hcp-properties}

\subsubsection{Preservation}\label{sec:hcp-preservation}
\input{lem-hcp-preservation-equiv}
\input{lem-hcp-preservation-subst}
\input{thm-hcp-preservation}

\subsubsection{Canonical Forms and Progress}\label{sec:hcp-canonical-forms-and-progress}

\paragraph{Canonical Forms}\label{sec:hcp-canonical-forms}
\input{def-hcp-canonical-forms}

\paragraph{Evaluation contexts}
\input{def-hcp-evaluation-contexts}
\input{def-hcp-evaluation-prefixes}

\paragraph{Progress}
\input{thm-hcp-progress}

\subsection{Relation to \cp}\label{sec:hcp2cp}

\subsubsection{\hcp shares \textsc{Mix}-free processes with \cp}
First off, we have the following theorems, relating terms with a singleton
hypersequent, i.e.\ \textsc{Mix}-free terms, in \hcp to terms in \cp and vice
versa.
\input{thm-cp2hcp-typing}
\input{thm-cp2hcp-equiv}
\input{thm-cp2hcp-reduction}
\input{thm-hcp2cp-typing}
\input{thm-hcp2cp-equiv}
\input{thm-hcp2cp-reduction}

\subsubsection{\hcp supports the same protocols as \cp}
In \hcp, we can use \textsc{H-Mix} to compose two independent processes, but
once composed, we cannot form any direct connections between the two.
Unconnected process can block on each other's communications, and behave
differently due to each other's choices.
Therefore, processes in \hcp are strictly more expressive than processes in \cp
and parallel compositions of processes in \cp.
One example of a process which is expressible in \hcp, but not in \cp, is
$$
\seq[{
  \cpCase{x}{\piPar{P}{Q}}{\piPar{P'}{Q'}}
}]{
  \Gamma \hsep \tmty{x}{A \with B}, \Delta
}
$$
where $\seq[P]{\Gamma}$, $\seq[P']{\Gamma}$, $\seq[Q]{\tmty{x}{A}, \Delta}$ and
$\seq[Q]{\tmty{x}{B}, \Delta}$.

However, we can prove that \hcp inhabits the same types as \cp. This is a
property of the logic, and we will present it as such, omitting the process
terms. We prove this by showing that the hypersequent separator can be
interpreted as a tensor.\footnote{%
  In most hypersequent systems, the hypersequent separator is interpreted as a
  disjunctive connective. However, here we interpret it as a conjunctive
  connective.}
First, however, we prove a lemma showing that we can interpret the structural
comma as a par.
\input{lem-hcp-bigparr}
\input{thm-hcp-bigtens}
As a corollary, we now know that \hcp inhabits the same types as \cp.
\input{fig-hcp-cc-hmix}


\section{Hypersequent Cyclic Classical Processes}\label{sec:hccp}

\subsection{Terms}\label{sec:hccp-term}
\input{def-hccp-terms}
\input{def-hccp-equiv}
\input{def-hccp-reduction}

\subsection{Types}\label{sec:hccp-types}
We keep the types, environments, and hyper-environments from \hcp, see
\cref{def:cp-types,def:cp-envs,def:hcp-hyperenvs}.
\input{def-hccp-typing}

\subsection{Properties}\label{sec:hccp-properties}

\subsubsection{Preservation}\label{sec:hccp-preservation}
\input{lem-hccp-preservation-equiv}
\input{lem-hccp-preservation-subst}
\input{thm-hccp-preservation}

\subsubsection{Canonical Forms and Progress}\label{sec:hccp-canonical-forms-and-progress}

\paragraph{Canonical Forms}\label{sec:hccp-canonical-forms}
\input{def-hccp-canonical-forms}

\paragraph{Evaluation Contexts}\label{sec:hccp-evaluation-contexts}
\input{def-hccp-evaluation-contexts}
\input{def-hccp-evaluation-prefixes}
\input{def-hccp-maximum-evaluation-prefixes}
\input{lem-hccp-maximum-evaluation-prefixes}

\paragraph{Progress}\label{sec:hccp-progress}
\input{thm-hccp-progress}

\subsection{Relation to \hcp}\label{sec:hcp2hccp}
\subsubsection{\hcp processes can be translated to \hccp}
The relation between \hccp and \hcp is slightly more complex than that between
\hcp and \cp.
In \cref{sec:hcp2cp}, we could simply insert terms in \cp into \hcp, as the term
language of \hcp is a simple extension of that of \cp.
In this case, however, we will have to start by defining a translation between
the two term languages.
\input{def-hcp2hccp-terms}
In the cases for $\tm{\cpCut{x}{P}{Q}}$, $\tm{\cpSend{x}{y}{P}{Q}}$ and
$\tm{\cpHalt{x}}$, while it does not look like much is happening, it is
important to realise that the constructs on the left-hand side are atomic,
whereas those on the right-hand side are complex. 

We can show that this translation respects typing.
\input{thm-hcp2hccp-typing}

We can show that this translation respects the structural congruence---that is
to say, if two terms are equivalent in \hcp, then their translations are
equivalent in \hccp.
\input{thm-hcp2hccp-equiv}

Furthermore, we can show that this translation respects reductions.
\input{thm-hcp2hccp-reduction}
\input{thm-hcp2hccp-reduction-inv}

\subsubsection{\hccp supports the same protocols as \hcp}\label{sec:hccp2hcp}
There is no neat translation from \hccp to \hcp, as there is no way to translate
processes of the form $\tm{\piSend{x}{}{P}}$ in \hcp---though there is probably
a way to simulate them.
Instead, similarly to \cref{sec:hcp2cp}, we will prove that \hccp inhabits the
same sequents as \hcp.
We prove this by showing that each proof in \hccp can be transformed into a
proof in \hcp.
The reduction behaviour between the two processes associated with these proofs,
however, may differ.
Therefore, we will present this as a property of the logic, and we will present
it as such, omitting the process terms.
\input{thm-hccp2hcp}
As a corollary, we now have that \hccp inhabits the same types as \cp.

\section{Priority-based Classical Processes}\label{sec:ccp}
The type system for \cp with cyclic communication structures (\ccp) is described
by \textcite{dardha2018}.
It is strictly more expressive than \cp and \hccp, as it allows cyclic
communication structures to be formed.
It enforces deadlock freedom by checking that the communication graph is
acyclic.
It uses the same definitions for terms as \hccp, and the following definitions
for structural congruence and term reduction. 
\input{def-ccp-equiv}
\input{def-ccp-reduction}
It uses the same types and environments as \hccp, though it annotates the
connectives of types with \emph{priorities}, e.g.\ as $\ty{A \tens^{\cs{o}} B}$.
A priority can be thought of as a label for an action.
Priorities form a partial order, where $\cs{o}<\cs{o'}$ means that the action
labeled with $\cs{o}$ must happen before the action labeled with $\cs{o'}$.
In \textcite{dardha2018}, priorities are natural numbers.
\input{def-ccp-pr}
\input{def-ccp-typing}

In this section, we will explore a revision of \ccp which has two benefits:
\begin{itemize}
\item
  we allow our priorities to be any partial order, instead of requiring them to
  be natural numbers;
\item
  and we generate a set of priority constraints during type checking, and return 
  this set as a result of type checking, instead of interleaving type and
  priority checking.
\end{itemize}
Why are these benefits?

The priority constraints are an encoding of (a sub-graph\footnote{%
  It is a sub-graph because some of the edges are enforced by the term syntax,
  and as such these are not checked by the priority constraints.
  For instance, in
  $\tm{\piBoundSend{x^{\cs{o}}}{y}{\piRecv{x^{\cs{o'}}}{z}{P}}}$,
  there is no reason to check that $\cs{o < o'}$, as there is no way for the
  receive action to take place before the send action.
} of) the communication graph, where $\cs{o < o'}$ means that the action
associated with $\cs{o}$ must happen before the action associated with
$\cs{o'}$.
In the case of parallel actions, we will have $\cs{o \parallel o'}$---i.e.\ %
there is \emph{no relation} between $\cs{o}$ and $\cs{o'}$. 
Natural numbers, however, are a total order, meaning that we will have to choose
either $\cs{o < o'}$, $\cs{o' < o}$, or $\cs{o = o'}$.
Therefore, by using priorities which have more structure than is required, we
lose some precision.

Gathering all priority constraints, instead of checking them during type
checking, makes it convenient to talk about
\begin{itemize}
\item 
  programs which are well-typed but (potentially) contain deadlocks;
\item
  programs which are deadlock free but also obey additional constraints on the
  ordering of actions, such as those that may arise from global types; 
\item
  translations between \ccp and other systems, such as \cp, because it separates
  the proofs of type preservation and that of constraint satisfaction.
\end{itemize}
We define two functions on priority-annotated types, in addition to the function
$\pr{\ty{A}}$ given in \cref{def:ccp-pr}:
\input{def-cccp-fg}
\input{def-cccp-un}
\input{def-cccp-typing}

\section{Safe Access Points}
\input{def-safeap-terms}
\input{def-safeap-equiv}
\input{def-safeap-reduction}
\input{def-safeap-types}
\input{def-safeap-typing}

\section{Global Types for Binary Sessions}\label{sec:globaltypes}
The work by \textcite{carbone2016} introduced globally-governed \cp (\gcp).
This work introduces a multiparty extension of \cp.
Processes in \gcp are ``typed'' by global types.
This is a bit of a misnomer, as global types are a whole host of things, but
types is not one of them, and it is probably best to see global types as the
arbiter processes to which they are compiled, together with some proof of
coherence.
With the addition of global types, \gcp adds several new---and in my eyes
orthogonal---features to \cp.

\begin{itemize}
\item
  \textbf{Global types stratify the global communication order}\\
  What do we mean by this? If there are two independent communications which can
  happen, say we have
  $\tm{({\piBoundSend{x}{y}{P}}\ppar{\piBoundSend{z}{w}{Q}}\ppar\dots)}$ and
  both $\tm{\piBoundSend{x}{y}{}}$ and $\tm{\piBoundSend{z}{w}{}}$ can happen,
  then we can use global types to stratify these communications. How? The
  potential for stratification in global types arises from the fact that we are
  basically writing an arbiter process. We are writing down who sends what to
  who when. For instance, for the example above, we could imagine some global
  type written as $\tm{{x}\to{\co{x}}.{z}\to{\co{z}}.\dots}$, where
  $\tm{\co{x}}$ and $\tm{\co{z}}$ are the endpoints on the other side of
  $\tm{x}$ and $\tm{z}$. This would require that the communication on $\tm{x}$
  happens before the communication on $\tm{z}$.
\item
  \textbf{Global types allow for broadcasting}\\
  There are two kinds of broadcasting in \gcp. First, the rules for $\tens\parr$
  and $\one\bot$ give the language some sort of ``gather'' operation and a
  wait-for-all operation. However, it is a well-known fact that the the
  constructs associated with the multiplicative connectives only perform
  ``plumbing''. Therefore, the gather construct that is implemented is not
  really gathering any information. Rather, it allows for some measure of
  abstraction in putting together sessions in which one process communicates
  with multiple other processes.

  The \emph{real} broadcasting operation is implemented by the $\plus\with$
  rule, which facilitates a single process sending one bit of information,
  identical in each case, to multiple recipients.

  The coherence rules presented by \textcite{carbone2016} are all asymmetric. In
  each case, \emph{one} process on the one side is communicating with multiple
  processes on the other. In each case, the side with \emph{one} process happens
  to be the ``strong'' side---the side which decides in which order
  communications happen $(\parr)$, and the side which makes the choice
  $(\plus)$. This, of course, is no coincidence. Instead, it has a much deeper
  connection to the logic. In each case, these stronger connectives are
  associated with \emph{unary} instead of binary rules. This makes it easy to
  expand a ``gather'' into a series of receive actions, and a choice into a
  series of choices.\footnote{This means that it should be possible to define a
    globally-governed variant of, say, \hccp or \ccp, in which both the plumbing
    and the choice are taken care of by broadcasting operations.}
\item
  \textbf{Global types allow for \emph{some} measure of non-determinism}\\
  If we see a broadcast operation as a series of binary sending operations, then
  we wouldn't expect those to be ordered among one another. The semantics for
  \gcp treat broadcast and gather as a \emph{single} operation, which only takes
  place when all intended recipients are ready. The semantics for \gcp, when
  compiled to \cp, however, imposes an ordering on the broadcasting (or
  gathering) process, by compiling the single action down to a series of
  actions. This is a bit odd.
\end{itemize}
We can easily model stratification of the global communication order in \cccp,
by simply making more demands of the final constraint set.
For instance, if we have a global type which requires
$\tm{{x}\to{\co{x}}.{z}\to{\co{z}}.\dots}$, i.e. that the first action on
$\tm{x}$ happens before the first action on $\tm{z}$, we can simply grab the
priorities associated with those actions---say $\cs{o}$ and $\cs{o'}$---and
require that the constraint $\cs{o} < \cs{o'}$ is present in the final
constraint set.

The broadcasting behaviours added by \gcp are\dots


\begin{definition}[Global types]\label{def:globaltypes}
  We define global types following \textcite{carbone2016}:
  \begin{gather*}
    \begin{array}{lcll}
      \tm{G}, \tm{H}
      & :=   & \gtyAx{x}{A}{y}                                         & \text{link} \\
      & \mid & \gtyTensParr{x}{y}{G}{H}                                & \text{multiparty gather} \\
      & \mid & \gtyOneBot{x}{y}                                        & \text{multiparty halt} \\
      & \mid & \gtyPlusWith{x}{y}{G}{H}                                & \text{multiparty choice} \\
      & \mid & \gtyNilTop{x}{y}                                        & \text{multiparty absurd}
    \end{array}
  \end{gather*}
\end{definition}

\begin{definition}[Coherence]\label{def:globaltypes-coherence}
  We define coherence following \textcite{carbone2016}:
  \begin{center} \gcpInfAx                     \end{center}\vspace{1\baselineskip}
  \begin{center} \gcpInfTensParr \gcpInfOneBot \end{center}
\end{definition}


\printbibliography

\end{document}
%%% Local Variables:
%%% TeX-master: "main"
%%% End:
