\begin{theorem}[Progress]\label{thm:hcp-progress}
  If $\seq[P]{\mathcal{G}}$, then there exists a $\tm{Q}$ such that either
  $\tm{P}\equiv\tm{Q}$ and $\tm{Q}$ is in canonical form, or $\reducesto{P}{Q}$.
\end{theorem}
\begin{proof}
  By induction on the structure of derivation for $\seq[{ P }]{ \mathcal{G} }$.
  The only interesting cases are when the last rule of the derivation is
  \textsc{Cut} or \textsc{H-Mix}. In every other case, the typing rule
  constructs a term which is in canonical form. 
  \\
  If the last rule in the derivation is \textsc{Cut} or \textsc{H-Mix}, we
  consider the maximum evaluation prefix $\tm{G}$ of $\tm{P}$, such that $\tm{P}
  = \tm{\cpPlug{G}{P_1 \dots P_{m+n+1}}}$ and each $\tm{P_i}$ is an action.
  The prefix $\tm{G}$ consists of $m$ mixes, $n$ cuts, and introduces $n$
  channels, but composes $m+n+1$ actions, at most $m+1$ of which are on the same
  side of all cuts.
  \begin{itemize}
  \item
    One of the actions composed by $\tm{G}$ is a link $\tm{\cpLink{x}{y}}$
    acting on a bound channel.
    Suppose that $\tm{x}$ is the bound channel.
    There exist evaluation contexts $\tm{E}$ and $\tm{E'}$ such that
    \[
      \tm{P} =
      \tm{\cpPlug{E}{\cpCut{x}{\cpPlug{E'}{\cpLink{x}{y}}}{R}}}.
    \]
    We rewrite by $\equiv$ to obtain
    \[
      \tm{P} =
      \tm{\cpPlug{E}{\cpPlug{E'}{\cpCut{x}{\cpLink{x}{y}}{R}}}}.
    \]
    We then reduce by applying \cpRedAxCut1.
    Similarly if $\tm{y}$ is the bound channel.
  \item
    Two of the actions composed by $\tm{G}$ are acting on the \emph{same} bound
    channel.
    Let the two actions be $\tm{P_i}$ and $\tm{P_j}$, and the bound channel
    $\tm{x}$.
    There exist evaluation contexts $\tm{E}$, $\tm{E_i}$, and $\tm{E_j}$ such
    that
    \[
      \tm{P} =
      \tm{\cpPlug{E}{\cpCut{x}{\cpPlug{E_i}{P_i}}{\cpPlug{E_j}{P_j}}}}
    \]
    We rewrite by $\equiv$ to obtain
    \[
      \tm{P} \equiv
      \tm{\cpPlug{E}{\cpPlug{E_i}{\cpPlug{E_j}{\cpCut{x}{P_i}{P_j}}}}}
    \]
    We then reduce by applying the appropriate \textbeta-reduction.
  \item
    Otherwise (at least) one of the processes acts on an external channel.
    No process $\tm{P_i}$ is a link acting on a bound channel.
    No two processes $\tm{P_i}$ and $\tm{P_j}$ act on the same channel $\tm{x}$.
    Therefore, $\tm{P}$ is equivalent to a process in canonical form.
  \end{itemize}
\end{proof}
%%% Local Variables:
%%% TeX-master: "main"
%%% End:
